Reassessing the role of theory and machine learning in forecasting civil conflict

### 2020-09-21 paper.Rmd
add cites
also move some footnotes and text discussion code lines to appendix

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves justified as easing interpretation. Smoothing is used to produce all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), for a total of 78 items, in order to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90%--17 of 19--clearly used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey.} (Both of the ambiguous cases include Blair as coauthor.) The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.  What is surprising is that anyone would choose the smoothed ROC implementation.

The next three issues we encountered all concern information in B&S Table 2.

@@

We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly, in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.

@@

This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consist of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.

The model specifications are also identical; i.e., they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing to forests of identical decision trees and, thus, identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined. We have documented details at https://github.com/andybega/Blair-Sambanis-replication/issues/5.]

@@

B&S present a confusion matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war onset, in the replication code, they are actually scored using incidence of civil war.

@@

```{r benefit-plot, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2)"}
table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
table2_benefit <- read_csv("data/table2-smooth-benefit.csv")

@@

@@

B\&S advocate the use of theory to guide prediction. But "theory" is not defined by B\&S, and may an ambiguous and undefined concept. `Theory'' has many interpretations in social science.  For some---the EITM movement, for example---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.  The additional statements can be subjected to empirical scrutiny.  An example might be William Riker's theory of coalitions which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically. A more recent example is found in The Political Logic of Survival.  For the purposes of discussion we call this strong theory.

However, for many,
a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables.  Typically it is fused together from a meta-analysis of prior work on the topic under consideration.   We call these types of theory weak theory.
They are weak because they do not include two characteristics of strong theory, though they may include one:
1. A theory is separate from the case or cases being examined. It is intented to be general.
2. A theory generates new statements that in principle can be empirically falsified, not proven.

@@

@@

>>>>

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation. However, smoothing is also used in all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90%--17 of 19--clearly used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey.} (Both of the ambiguous cases include Blair as coauthor.) The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.  What is surprising is that anyone would choose the smoothed ROC implementation.

The next three issues we encountered all concern information in B&S Table 2, which is the key table in which B&S report AUC-ROC values to demonstrate the "Escalation" models superiority.

@@

We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set. This appears to be an easily corrected coding error on the part of B&S.

@@

This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consist of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split. 
The model specifications are also identical; i.e., they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing to forests of identical decision trees and, thus, identical predictions.

@@

B&S present a confusion matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war _onset_, in the replication code, they are actually scored using _incidence_ of civil war. By definition the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict. When forecasting rare events, it is common to have many false positives, i.e. instances where we predict onset but not onsets occur. _Ex ante_, using incidence rather than onset to score onset forecasts should decrease the number of false positives and correspondingly increase the number of true positives (good predictions) and false negatives (missed onsets).

@@

```{r benefit-plot, fig.height=3.5, fig.width=5.5, out.width = ".8\\linewidth", fig.align = "center", fig.cap = "Gain from using smoothed ROC to calculate AUC, for each model reported in Tables 3 and 4 (B\\&S Tables 1 and 2). The Escalation model is the only model which consistently benefits from using smoothed ROC curves, and this advantage accounts for its apparent superiority to the other models."}
table1_benefit <- read_csv("data/table1-smooth-benefit.csv") 
table2_benefit <- read_csv("data/table2-smooth-benefit.csv")

@@

Predicting rare events like civil war onset is exceedingly difficult, and it is not unusual at all that most of the predictions are false positives. This will by definition be the case when there are no onsets in a time period, as is true with the default "Assuming Persistence" coding that is in the data. Correctly predicting the two onsets in the alternative "Assuming Change" coding is at face value a credit to the model's accuracy.

@@

B\&S advocate the use of theory to guide prediction. But "theory" is not defined by B\&S, and may an ambiguous and undefined concept. `Theory'' has many interpretations in social science.  For some---the EITM movement, for example [@aldrich:etal:2008]---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.  The additional statements can be subjected to empirical scrutiny.  An example might be William Riker's theory of coalitions [@riker:1962] which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically. A more recent example is found in The Political Logic of Survival [@bdm:etal:2005]. For the purposes of discussion we call this strong theory.

However, for many, a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables. Typically it is fused together from a meta-analysis of prior work on the topic under consideration. We call these types of theory weak theory.
They are weak because they do not include two characteristics of strong theory, though they may include one:
1. A theory is separate from the case or cases being examined. It is intended to be general.
2. A theory generates new statements that in principle can be empirically falsified, not proven.

@@

## Code references for coding errors

### Section 3.2: Incorrect “Weighted by PITF” Implementation

The intention with the "Weighted by PITF" model is to combine the PITF test predictions with the Escalation test predictions, i.e. weight the latter with the former. Due to an error, the PITF _training_ set predictions are used, not the _test_ set predictions.

See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly, in the 6-month version of this file.

Additional details in issue #3 in the GitHub repo.

### Section 3.3: Incorrect “PITF Split Population” Implementation

The intention is to train seperate random forests on data split into high and low-risk groups based on PITF predictions, and then combine them back into one random forest. Actually both random forests are trained on identical data, essentially production a random forest that just has N_trees \* 2.

Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined. Details are documented in issue #5 in the GitHub repo.

### Section 3.5: Incorrect scoring of the 2016 forecasts

@@

Only the "binormal" smoothing method produces a pattern of AUC-ROCs that clearly elevates the Escalation model above the alternatives. The "density" method does not on average change the AUC-ROC values very much from their empirical ROC curve equivalents. The "fitdistr" method tends to increase all AUC-ROC values, although maybe slightly more so for the Escalation model than the main alternatives.
